Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2022-05-27 methodology_pnorm.tex
updates
pgpareto github

%\begin{figure}[ht]
%    \centering
%    \includegraphics{images/p_sphere}
%    \caption{Unit hyperspheres established on $\mathcal{L}_p$, for $p = 1,\ldots,10$, for two dimensions.\label{fig:psphere}}
%\end{figure}

Consider a vector $\bm{s} \in {\mathbb R}^d$, and let the $\mathcal{L}_p$-norm be defined as
  \begin{equation*}
    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
  \end{equation*}
for $p>0$. The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively. 
The $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
  \begin{equation*}
    \lVert \bm{s} \rVert_{\infty}
      = \lim\limits_{p\to\infty} \lVert \bm{s} \rVert_p
      = \max_{\ell\in\lbrace1,\ldots,d\rbrace}s_{\ell}.
  \end{equation*}
To define an angular measure,  we are interested in the direction of
vectors in ${\mathbb R}_{+}^d$, thus, we project them onto
  \begin{equation*}
    {\mathbb S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
  \end{equation*}
  the $\mathcal{L}_p$-norm unit hypersphere.
  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto ${\mathbb S}_{p}^{d-1}$ is given as
  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in {\mathbb S}_{p}^{d-1}$.
  Observations on one hypersphere can be projected without loss of information onto another by
  dividing by the defining norm of the target hypersphere. In fact, letting 
  \begin{equation*}
    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}.
  \end{equation*}
  the transformation
  \begin{equation}
    \label{eqn:pnormt}
    T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
  \end{equation}
  is invertible with
  \begin{equation}
    \label{eqn:pnormtinv}
    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right) =
      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
  \end{equation}
  The Jacobian of the transformation takes the form
  \begin{equation}
    \label{eqn:pnormjac}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
        {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^p_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$. Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable. Thus, even though the projected distribution on ${\mathbb S}_\infty^{d-1}$ is defined, there is no corresponding density.

% EOF

>>>>

Consider a vector $\bm{s} \in {\mathbb R}^d$, and let the $\mathcal{L}_p$-norm be defined as
  \begin{equation*}
    \lVert \bm{s} \rVert_p = \left({\textstyle\sum}_{\ell = 1}^d \lvert s_{\ell}\rvert^p\right)^{\frac{1}{p}},
  \end{equation*}
for $p>0$. The absolute and Euclidean norms are obtained for $p=1$ and $p=2$ respectively. 
The $\mathcal{L}_{\infty}$ norm can be obtained as a limit, 
  \begin{equation*}
    \lVert \bm{s} \rVert_{\infty}
      = \lim\limits_{p\to\infty} \lVert \bm{s} \rVert_p
      = \underset{\ell = 1}{\overset{d}{\vee}}s_{\ell}.
  \end{equation*}
To define an angular measure,  we are interested in the direction of
vectors in ${\mathbb R}_{+}^d$, thus, we project them onto
  \begin{equation*}
    {\mathbb S}_{p}^{d-1} = \left\lbrace \bm{s} : \bm{s} \in {\mathbb R}_{+}^{d}, \lVert \bm{s}\rVert_{p} = 1\right\rbrace,
  \end{equation*}
  the $\mathcal{L}_p$-norm unit hypersphere.
  Given $\bm{x}\in {\mathbb R}^d_+$ the projection onto ${\mathbb S}_{p}^{d-1}$ is given as
  $\bm{y} = \bm{x} / \lVert \bm{x}\rVert_p \in {\mathbb S}_{p}^{d-1}$.
  Observations on one hypersphere can be projected without loss of information onto another by
  dividing by the defining norm of the target hypersphere. In fact, for finite $p$, letting 
  \begin{equation*}
    y_d = \left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}},
  \end{equation*}
  the transformation
  \begin{equation}
    \label{eqn:pnormt}
    T(x_1,\ldots,x_d) = \left(\pnorm{\bm{x}}{p}, \frac{x_1}{\pnorm{\bm{x}}{p}},
                          \ldots , \frac{x_{d-1}}{\pnorm{\bm{x}}{p}}\right) = (r,y_1,\ldots,y_{d-1})
  \end{equation}
  is invertible with
  \begin{equation}
    \label{eqn:pnormtinv}
    T^{-1}\left(r,y_1,\ldots,y_{d-1}\right) =
      \left(ry_1,\ldots,ry_{d-1}, r\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}}\right).
  \end{equation}
  The Jacobian of the transformation takes the form
  \begin{equation}
    \label{eqn:pnormjac}
    r^{d-1}\left[\left(1 - {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\right)^{\frac{1}{p}} +
        {\textstyle\sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(1 - {\textstyle\sum}_{l=1}^{d-1} y_{\ell}^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) can be used to define a density on ${\mathbb R}^d_+$, and then obtain a density on ${\mathbb S}_{p}^{d-1}$, for any finite $p>0$. 
  %Unfortunately in the limit, when $p\rightarrow +\infty$, the transformation is not differentiable. Thus, even though the projected distribution on ${\mathbb S}_\infty^{d-1}$ is defined, there is no corresponding density \temp{when the distribution is projected in this manner.  However, for large but finite $p$, $\mathbb{S}_p^{d-1}$ will tend to approach $\mathbb{S}_{\infty}^{d-1}$}.

%   \temp{
%   We might also ask what degree of distortion in the distribution is imposed by the difference in projections
%   from $\mathbb{S}_{\infty}^{d-1}$ to $\mathbb{S}_p^{d-1}$.  Ideally, we might consider the ratio of surface areas between the two surfaces. As there is no general closed form expression for the surface area of a unit $n$-sphere under the $\mathcal{L}_p$ norm, it is difficult to assess in this manner.  There is, however, a closed-form expression for the \emph{volume} of the positive orthant of the unit $n$-sphere under the $\mathcal{L}_p$ norm.  This is not directly comparable, however, the closer the volume is to 1, the closer ${\mathbb S}_{p}^{d-1}$ is to ${\mathbb S}_{\infty}^{d-1}$.
%     \[
%         \text{Volume}~\left({\mathbb S}_{p}^{d-1}\right) = \frac{\Gamma\left(\frac{1}{p} + 1\right)^{d}}{\Gamma\left(\frac{d}{p} + 1\right)}
%     \]
%     }
% EOF

### 2022-05-27 methodology_projgamma.tex
pgpareto github

\subsection{Projected gamma family\label{subsec:projgamma}}
  A natural  distribution to consider in ${\mathbb R}^p_+$ is given by a product of independent
  univariate Gamma distributions. Let
    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
  \begin{equation}
  \label{joint}
    f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
      \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]
      \times r^{d-1}\left[y_d +
            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Integrating out $r$ yields the resulting \emph{Projected Gamma} density
  \begin{equation}
    \label{eqn:projgamma}
    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) =
          \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
                y_{\ell}^{\alpha_{\ell} - 1}\right]
      \times \left[y_d +
          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
      \times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}} ,
  \end{equation}
  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any $p>0$. To avoid identifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
  \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
  for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.
  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.

  The projected gamma family is simple to specify and has very tractable computational properties. Thus, we use it as a building block for the angular measure $\Phi$ models. To build a flexible family of distributions in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
  \begin{equation} \label{eqn:PGmix}     
     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
  \end{equation} 
  where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ is drawn from a random measure. In particular, assuming a Dirichlet process prior, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
  \begin{equation}\label{eqn:dppg}    \bm{y}_i \sim \text{PG}(\bm{y}_i|\bm{\theta}_i) , \;\;\; \bm{\theta}_i \sim G, \;\;\; G \sim \mathcal{DP}(\eta, G_0),
  \end{equation}
  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.

  Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
  For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section.

@@

\subsubsection{Inference for the projected Gamma mixture model}
The Dirichlet process mixture model groups observations into stochastically assigned clusters.  Building out the methods of inference for Equation~\ref{eqn:dppg}, let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be the number of extant 
% {\bf (remaining?)} observations
clusters, not including any singleton that observation $i$ is in. Under this model, the probability of cluster membership for a given observation is proportional to
\begin{equation*}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
If $G_0$ is not a conjugate prior for the kernel, the integral in the above formula may not be
available in closed form. Using Algorithm 8 from \cite{neal2000}, by Monte Carlo integration, drawing  $m$ candidate
clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where
the probability of cluster membership is proportional to
\begin{equation}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
If a candidate cluster is selected,
%{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
then $\delta_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.

A key feature of the the projected Gamma distribution is its computational properties. We augment  $\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, for each observation. Using Equation \eqref{joint} we observe that the 
full conditional of $r_i$ is easy to sample from, as it is given as
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(\sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
\end{equation}
Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
\end{equation}
We first consider a centering distribution given by a product of independent Gammas:
\begin{equation}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
\end{equation}
This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  We refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\zeta_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Samples of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
The full conditional for $\beta$ is 
\begin{equation}
    \label{eqn:betafc}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\mid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.

>>>>

\subsection{Projected gamma family\label{subsec:projgamma}}
  A natural  distribution to consider in ${\mathbb R}^d_+$ is given by a product of independent
  univariate Gamma distributions. Let
    $\bm{ X} \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{\ell}\mid\alpha_{\ell},\beta_{\ell}\right)$, 
    $\alpha_\ell$ and $\beta_\ell$ are the shape and scale parameters, respectively. Using Equations (\ref{eqn:pnormt}) -- (\ref{eqn:pnormjac}) we have the joint density
  \begin{equation}
  \label{joint}
    f(r,\bm{ y}) = \prod_{\ell = 1}^{d}
      \left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}(ry_{\ell})^{\alpha_{\ell} - 1}
          \exp\lbrace-\beta_{\ell}ry_{\ell}\rbrace\right]
      \times r^{d-1}\left[y_d +
            {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right].
  \end{equation}
  Integrating out $r$ yields the resulting \emph{Projected Gamma} density
  \begin{equation}
    \label{eqn:projgamma}
    \text{PG}(\bm{ y}\mid\bm{ \alpha},\bm{ \beta}) =
          \prod_{\ell = 1}^d\left[\frac{\beta_{\ell}^{\alpha_{\ell}}}{\Gamma(\alpha_{\ell})}
                y_{\ell}^{\alpha_{\ell} - 1}\right]
      \times \left[y_d +
          {\textstyle \sum}_{\ell = 1}^{d-1}y_{\ell}^p\left(y_d^p\right)^{\frac{1}{p} - 1}\right]
      \times \frac{\Gamma({\textstyle\sum}_{\ell = 1}^d\alpha_{\ell})}{\left({\textstyle\sum}_{\ell = 1}^d
                    \beta_{\ell}y_{\ell}\right)^{{\scriptstyle\sum_{\ell = 1}^d \alpha_{\ell}}}} ,
  \end{equation}
  defined for $\bm{y}\in {\mathbb S}_p^{d-1}$, and for any $p>0$. To avoid identifiability problems when estimating the shape and scale parameters, we set $\beta_1 = 1$.
  \cite{nunez2019} obtain the density in Equation (\ref{eqn:projgamma})
  for $p=2$ as a multivariate distribution for directional data, using spherical coordinates.  
  For $\bm{ y}\in {\mathbb S}_1^{d-1}$, and $\beta_{\ell} = \beta$ for all $\ell$, the density in Equation (\ref{eqn:projgamma}) corresponds to that of a Dirichlet distribution.

  The projected gamma family is simple to specify and has very tractable computational properties. Thus, we use it as a building block for the angular measure $\Phi$ models. To build a flexible family of distributions in ${\mathbb S}_p^{d-1}$ we consider mixtures of projected gamma densities defined as
  \begin{equation} \label{eqn:PGmix}     
     f(\bm{y}) = \int_\Theta \mathcal{PG}(\bm{y}|\bm{\theta}) dG(\bm{\theta}), 
  \end{equation} 
  where $\bm{\theta} = (\bm{\alpha}, \bm{\beta})$. Following a Bayesian non-parametric approach \citep{Ferguson74,Antoniak1974} we assume that $G$ is drawn from a random measure. In particular, assuming a Dirichlet process prior for $G$, we have a hierarchical formulation of the mixture model that, for a vector of observations $\bm{y}_i$, is given by
  \begin{equation}\label{eqn:dppg}    \bm{y}_i \sim \text{PG}(\bm{y}_i|\bm{\theta}_i) , \;\;\; \bm{\theta}_i \sim G, \;\;\; G \sim \mathcal{DP}(\eta, G_0),
  \end{equation}
  where $\mathcal{DP}$ denotes a Dirichlet process, $\eta$ is the precision parameter, and $G_0$ is the centering distribution.

  Unfortunately, in the limit when $p\rightarrow \infty$, the 
  projection transformation is not differentiable. Thus, a simple
  formula like Equation \ref{eqn:projgamma} for the
  projected gamma density on ${\mathbb S}_\infty^{d-1}$ is not 
  available. Our strategy consists of describing the angular distribution $\Phi$ using a sample based approach with the following steps: (i) Apply the transformation in Equation (\ref{eqn:standardization}) to the original data; (ii) Obtain the subsample of the standardized observations that satisfy $R>1$; (iii) Take a finite $p$ and project the observations onto ${\mathbb S}_p^{d-1}$; (iv) Fit the model in Equation (\ref{eqn:PGmix}) to the resulting data and obtain samples from the fitted model; (v) Project the resulting samples onto ${\mathbb S}_\infty^{d-1}$.
  For step (iv) we use a Bayesian approach that is implemented using a purposely developed Markov chain Monte Carlo that is described in the next section.

@@

\subsubsection{Inference for the projected Gamma mixture model}
In a sample-based inference approach, for a given iteration the Dirichlet process mixture model groups observations
    into stochastically assigned clusters, where members of a cluster share distributional parameters. Building out 
    the methods of inference for Equation~\ref{eqn:dppg}, 
    let $n_j^{-i}$ be the number of observations in cluster $j$ not including observation $i$.  Let $J^{-i}$ be 
    the number of extant clusters, not including any singleton that observation $i$ is in. Under this model, 
    the probability of cluster membership for a given observation is proportional to
\begin{equation*}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \eta\int\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)dG_0(\bm{\alpha}_j,\bm{\beta}_j)\hspace{0.5cm} &\text{~for~}j = J^{-i} + 1.
        \end{cases}
\end{equation*}
Note the top case is iterating over extant clusters, and the bottom case is for a \emph{new} cluster. 
If $G_0$ is not a conjugate prior for the kernel density, the integral in the above formula may not be
available in closed form. We solve this by using Algorithm 8 from \cite{neal2000}: by Monte Carlo integration, 
we draw $m$ candidate clusters, $\bm{\alpha}_j,\bm{\beta}_j$ for $j = J^{-i} + 1,\ldots, J^{-i} + m$ from $G_0$. 
Then, we sample the cluster indicator $\delta_i$ from extant or candidate clusters, where
the probability of cluster membership is proportional to
\begin{equation}
    \text{Pr}\left[\delta_i = j\mid\ldots\right] \propto \begin{cases}
        n_j^{-i}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace 1,\ldots,J^{-i}\rbrace\\
        \frac{\eta}{m}\mathcal{PG}\left(\bm{y}_i\mid\bm{\alpha}_j,\bm{\beta}_j\right)\hspace{0.5cm} &\text{~for~}j \in \lbrace J^{-i} + 1,\ldots, J^{-i} + m\rbrace.
        \end{cases}
\end{equation}
If a candidate cluster is selected,
%{\bf you first have to say that you sample from this distribution. What is the ``candidate cluster''?}
then $\delta_i = J = J^{- i} + 1$, and the associated cluster parameters are saved.

A key feature of the the projected Gamma distribution is its computational properties. We augment 
$\mathcal{PG}(\bm{y}_i\mid\bm{\alpha}_i,\bm{\beta}_i) $ by introducing a latent radial component $r_i$, 
for each observation. Using Equation \eqref{joint} we observe that the 
full conditional of $r_i$ is easy to sample from, as it is given as
\begin{equation}
    r_i\mid\bm{\alpha}_i,\bm{\beta}_i \sim \mathcal{G}\left(r_i \bigmid \sum_{\ell = 1}^d\alpha_{i\ell}, \sum_{\ell = 1}^d\beta_{\ell} y_{i\ell}\right).
\end{equation}
Moreover,  the full conditional for $\bm{\alpha}_j,\bm{\beta}_j$ is then proportional to
\begin{equation}
    f(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{Y},\bm{r},\bm{\delta},\ldots) \propto \prod_{i:\gamma_i = j}\prod_{\ell = 1}^d\mathcal{G}\left(r_iy_{i\ell}\mid\alpha_{j\ell},\beta_{j\ell}\right) \times dG_0(\bm{\alpha}_j,\bm{\beta}_j).
\end{equation}
Note that the ordering of the products can be reversed in the likelihood, indicating that given $\bm{r}$, 
the likelihood becomes separable by dimension.
We first consider a centering distribution given by a product of independent Gammas:
\begin{equation}
    G_0(\bm{\alpha}_j,\bm{\beta}_j\mid \bm{\xi},\bm{\tau},\bm{\zeta},\bm{\sigma}) = \prod_{\ell = 1}^d\mathcal{G}(\alpha_{j\ell}\mid \xi_{\ell},\tau_{\ell})\times\prod_{\ell = 2}^d\mathcal{G}(\beta_{j\ell}\mid\zeta_{\ell},\sigma_{\ell}).
\end{equation}
This model is completed with independent Gamma priors on $\xi_{\ell}$, $\tau_{\ell}$, $\zeta_{\ell}$, $\sigma_{\ell}$.  We also assume a Gamma prior on $\eta$, that is updated via the procedure outlined in \cite{escobar1995}.  We refer to this model as the \emph{projected Gamma--Gamma} (PG--G) model.  An advantage of the PG--G model is that, thanks to conjugacy, the rate parameters can easily be integrated out for inference on $\bm{\alpha}_j$.  Then, the full conditional for $\alpha_{j\ell}$ takes the form
\begin{equation}
    \pi(\alpha_{j\ell}\mid \bm{r},\bm{Y},\bm{\gamma},\xi_\ell,\tau_\ell,\zeta_\ell,\sigma_\ell) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i\ell}\right)^{\alpha_{j\ell} - 1}\alpha_{j\ell}^{\xi_\ell - 1}e^{-\tau_\ell \alpha_{j\ell}}\times\Gamma\left(n_j\alpha_{j\ell} + \zeta_{\ell}\right)}{\Gamma^{n_j}(\alpha_{j\ell})\left(\sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right)^{n_j\alpha_{j\ell} + \zeta_{\ell}}}
\end{equation}
for $\ell = 2,\ldots,d$.  For $\ell = 1$, as $\beta_{1} := 1$, the full conditional takes the simpler form
\begin{equation}
    \label{eqn:alpha1update}
    \pi(\alpha_{j1}\mid\bm{r},\bm{Y},\bm{\gamma},\xi_1,\tau_1) \propto \frac{\left(\prod_{i:\gamma_i = j}r_iy_{i1}\right)^{\alpha_{j1} - 1}\alpha_{j1}^{\xi_1 - 1}e^{-\tau_1\alpha_{j1}}}{\Gamma^{n_j}(\alpha_{j1})}.
\end{equation}
Samples of $\alpha_{j\ell}$ can thus be obtained using a Metropolis step. In our analysis, we first transform $\alpha_{j\ell}$ to the log scale, and use a normal proposal density.
The full conditional for $\beta$ is 
\begin{equation}
    \label{eqn:betafc}
    \beta_{j\ell}\mid\bm{r},\bm{Y},\alpha,\zeta_{\ell},\sigma_{\ell} \sim \mathcal{G}\left(\beta_{j\ell}\bigmid n_j\alpha_{j\ell} + \zeta_\ell, \sum_{i:\gamma_i = j}r_iy_{i\ell} + \sigma_{\ell}\right),
\end{equation}
for $\ell = 2,\ldots, d$.  Updating $\beta_{j\ell}$ is done via a Gibbs step.  The hyper-parameters $\xi_{\ell},\tau_{\ell},\zeta_{\ell},\sigma_{\ell}$ follow similar Gamma-Gamma update relationships.  We also explore a restricted form of this model, where $\beta_{\ell} = 1$ for all $\ell$.  Under this model, we use the full conditional in Equation~\ref{eqn:alpha1update} for all $\ell$, and omit inference on $\bm{\zeta},\bm{\sigma}$.  We refer to this model as the \emph{projected restricted Gamma--Gamma} (PRG--G) model.
